---
title: "Creating a Merged Person-Household ACS PUMS Dataset"
author: "Data Team, [Center on Budget and Policy Priorities](https://www.cbpp.org)"
date: "Spring 2021"
output: 
  html_document:
    theme: paper
    toc: true
    toc_depth: 1
---

```{r initialize-statamarkdown, echo=FALSE, message=FALSE}
library(Statamarkdown)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(collectcode=TRUE, 
                      # comment out the line below to execute Stata code
                      # eval=FALSE, 
                      savedo=TRUE)
```

```{stata render-setup, include=FALSE, savedo=FALSE}
set rmsg off
set more off
set linesize 120
```

This tutorial will demonstrate how to make a merged person-household ACS PUMS dataset and how to create and use household-level variables based on the characteristics of persons within the household. We'll use several commands from the [`cbppstatautils` Stata package](https://github.com/CenterOnBudget/cbpp-stata-utils), so please install the package if you haven't already.

```{stata, include=FALSE, collectcode=FALSE}
* Check if cbppstatautils package installed, and if not, install it
capture which cbppstatautils
if _rc != 0 {
	net install cbppstatautils, from("https://raw.githubusercontent.com/CenterOnBudget/cbpp-stata-utils/master/src") replace
}
```


# Creating a merged person-household dataset

First, download both person and household records.

```{stata, echo=-1, eval=FALSE, collectcode=FALSE}
* Download both record types' datasets
get_acs_pums, year(2019) sample(1) record_type("both") state("vt")      ///
              keep_zip replace
```

```{stata, include=FALSE, savedo=FALSE, collectcode=FALSE}
capture confirm file "acs_pums/2019/1_yr/psam_h50.dta"
if _rc != 0 {
  get_acs_pums, year(2019) sample(1) state("vt") record_type("both") keep_zip
}
```

Merge the household and person records together on `serialno`, the household record ID.

```{stata, echo=-1}
* Merge the records together by serialno, the household record ID
use "acs_pums/2019/1_yr/psam_h50.dta", clear
merge 1:m serialno using "acs_pums/2019/1_yr/psam_p50.dta"
```

All household records observations that didn't match to any person records (`_merge == 1`) should be vacant housing units `(np == 0`).

```{stata, echo=-1}
* Check that unmatched records are vacant housing units
count if _merge == 1
count if np == 0
```

We'll do some more prep: label the dataset, and adjust dollar-denominated variables for inflation.

```{stata, echo=-1}
* Some dataset prep
label_acs_pums, year(2019) sample(1)
generate_acs_adj_vars
```

And since here we are interested in households and people, we'll drop vacant housing units — household records where np, the number of people in the household, is zero.

```{stata, echo=-1}
* Drop vacant housing units
drop if np == 0
```

Let's check out the merged dataset. We'll preview a mix of household- and person-level variables:

- `serialno` is the household ID variable.
- `ten` is a household-level variable indicting household tenure (homeowner, renter, etc.).
- `sporder` is the person ID variable.
- `rc` is a person-level variable indicating whether the person is a child related to the householder.

```{stata, echo=FALSE, savedo=FALSE}
sort type serialno sporder
list serialno ten sporder rc in 1/15, separator(0) noobs
```


# Tabulating person-level variables by household-level variables

Set the survey design using person weights since we will be tabulating person-level variables.

```{stata, echo=c(-1:-3)}
* Tabulating person-level variables by household-level variables

* Set the survey design using person weights
svyset_acs, record_type("person")
```

We can tabulate person-level variables by household-level variables.

Number of related children by tenure of household:
```{stata, echo=-1}
* Related children by household tenure
svy: total rc, over(ten)
```

Disability status by household SNAP receipt:
```{stata, echo=-1}
* Disability status by household SNAP receipt
svy: proportion dis, over(fs)
```


# Generating and tabulating new household-level variables 

We can create new household-level variables based on the characteristics of persons within the household, such as a variable indicating whether anyone in the household is a child attending school.

```{stata, echo=c(-1:-3), results="hide"}
* Creating and tabulating new household-level variables

* Variable indicating whether anyone in household is a child attending school
generate schkid = agep < 18 & inlist(sch, 2, 3) if !missing(sch)
egen hh_has_schkid = max(schkid), by(serialno) 
```

Recall that our dataset has one observation per person. So tabulating hh_has_schkid now will count the number of _people in households_ with a child attending school, and we must use person weights since we are tabulating people. 

```{stata, echo=-1}
* People in households with a child attending school
svy: total hh_has_schkid
```

If we want the number of _households_ with a child attending school, we need to do two things:

  1. Use household weights, since we are tabulating households
  2. Limit our sample to one observation per household. The easiest way to do that is to keep only the householder's record, `relshipp == 20`.[^1] 
  
```{stata, echo=-1}
* Households with a child attending school
svyset_acs, record_type("household")
svy: total hh_has_schkid if relshipp == 20
```

If you are switching back and forth between person-level and household-level tabulations in your merged file, and need to check whether the survey settings are currently set for household or person weights, you can run the `svyset` command without arguments to display the current survey settings. Recall that person weight variables start with "pwgtp" and household weight variables start with "wgtp."

```{stata, echo=-1}
* Check current survey settings
svyset
```


[^1]: The household relationship variable is `relshipp` in the 2019 sample, `relp` in the 2010-2018 samples, and `rel` in prior samples. The values of `relshipp` are different from previous years' versions of the variable.
